---
title: 'Assignment 2: Data visualization'
author: "Tamas Nagy"
output: html_document
editor_options: 
  chunk_output_type: console
---

You will have to create 3 plots based on the datasets and instructions detailed below. You will find the plots themeselves in the `assignments/assignment_2_plots`. Your task is to write the code that will reproduce the plots as closely as possible.

# Skills needed to solve this assignment

-   Using R and RStudio, reading data
-   Reporting using RMarkdown
-   Using Git and Github (for submitting the task)
-   Data manipulation (e.g. dplyr, tidyr), and working with factors (forcats)
-   Data visuzlization (ggplot2)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(tidytuesdayR)
```

## Task 1: Climbing expeditions

The 2020-09-22 TidyTueday datasets are about climbing expeditions. From the three datasets, use the "expeditions". Reproduce the plot below! Notice a few things:

-   Use `forcats::fct_lump()` to get the 15 most frequent peaks, and drop the "Other" category.
-   The bars are ordered by the sum of all expeditions (use `fct_reorder()`).
-   The bar colors use the viridis palette and light theme.

```{r}

install.packages("forcats")
library(forcats)
library(dplyr)

expeditions_filtered <- expeditions %>%
  mutate(top_15 = fct_lump(peak_name, n = 15, other_level = "Other")) %>%
  filter(top_15 != "Other")

View(expeditions_filtered)
library(dplyr)

expeditions %>%
  count(peak_name, sort = TRUE) %>%
  slice_head(n = 15)

library(dplyr)
library(forcats)
library(ggplot2)

top_peaks <- expeditions %>%
  count(peak_name, sort = TRUE) %>%
  slice_head(n = 15) %>%
  mutate(peak_name = fct_reorder(peak_name, n))

ggplot(top_peaks, aes(x = peak_name, y = n)) +
  geom_col() +
  coord_flip() +
  labs(x = "Peak", y = "Number of expeditions", title = "Top 15 Most Frequent Peaks")

library(viridis)
top_peaks <- expeditions %>%
  count(peak_name, sort = TRUE) %>%
  slice_head(n = 15) %>%
  mutate(peak_name = fct_reorder(peak_name, n))

ggplot(top_peaks, aes(x = peak_name, y = n, fill = n)) +
  geom_col() +
  scale_fill_viridis_c() +
  coord_flip() +
  theme_light() +
  labs(x = "Peak", y = "Number of expeditions", title = "Top 15 Most Frequent Peaks")


```

## Task 2: PhDs awarded

The 2019-02-19 TidyTueday dataset is about phd-s awarded by year and field. There is only one dataset, it is called `phd_by_field`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all phd-s by broad fields.
-   To make the x axis breaks pretty, use `scales::pretty_breaks()`, to make the y axis labels comma formatted, use `scales::comma_format()`.
-   The line size is 1.2, the colors are from the brewer "Dark2" palette. The theme is set to minimal.

```{r}
install.packages("readr")
library(readr)
url <- "https://raw.githubusercontent.com/rfordatascience/tidytuesday/refs/heads/main/data/2019/2019-02-19/phd_by_field.csv"
phd_data <- read_csv(url)
View(phd_data)

library(dplyr)
phd_counts <- phd_data %>%
  group_by(broad_field) %>%
  summarise(total_phds = n()) %>%
  arrange(desc(total_phds))
phd_counts

library(ggplot2)
library(scales)
ggplot(phd_counts, aes(x = broad_field, y = total_phds)) +
  geom_col(fill = "steelblue") +
  scale_y_continuous(labels = comma_format()) +
  scale_x_discrete(breaks = scales::pretty_breaks()) +  
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

library(RColorBrewer)
ggplot(phd_counts, aes(x = broad_field, y = total_phds, fill = broad_field)) +
  geom_col(color = "black", size = 1.2) +
  scale_fill_brewer(palette = "Dark2") +
  scale_y_continuous(labels = comma_format()) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none")

```

## Task 3: Commute in the US

The 2019-11-05 TidyTueday dataset is about commuting to work in each city in the US by bike or on foot. There is only one dataset, it is called `commute`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all commutes by state.
-   Both axis scales are log transformed and the labels comma formatted, using `scales::comma_format()`
-   The point size is 2, . The theme is set to light.

```{r}
library(readr)

commute_mode <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2019/2019-11-05/commute.csv")

View(commute_mode)

library(dplyr)

commute_by_state <- commute_mode %>%
  group_by(state) %>%
  summarise(total_commutes = sum(n, na.rm = TRUE))

View(commute_by_state)

library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)

commute_wide <- commute_mode %>%
  filter(mode %in% c("Bike", "Walk")) %>%
  group_by(state, mode) %>%
  summarise(total = sum(n, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = mode, values_from = total)

ggplot(commute_wide, aes(x = Bike, y = Walk)) +
  geom_point(color = "steelblue", size = 2) +
  geom_text(aes(label = substr(state, 1, 2)), color = "white", size = 3) +
  scale_x_log10(labels = comma_format()) +
  scale_y_log10(labels = comma_format()) +
  labs(x = "Bike commuters", y = "Walk commuters", title = "Commuting by Bike vs. Walk (per state)") +
  theme_light()



```
